Microxanox: an R package for simulating an aquatic \(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.

### 2022-05-26 inst/manuscript/manuscript_microxanox.tex
work on ms

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

@@

\usepackage{lineno} % add
  \linenumbers % turns line numbering on

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

%% \documentclass[preprint,12pt, a4paper]{elsarticle}

@@

\begin{document}

\begin{frontmatter}

  \title{Microxanox: an R package for simulating an aquatic
\(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
    \author[University of Zürich]{Rainer M Krug%
  %
  \fnref{Corresponding Author}}
   \ead{Rainer.Krug@uzh.ch; Rainer@krugs.de} 
    \author[University of Zürich]{Owen L. Petchey}
   \ead{Owen.Petchey@ieu.uzh.ch} 
      \affiliation[University of Zürich]{Department of Evolutionary
Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{Corresponding Author}
    \fntext[2]{Equal contribution}

  \begin{abstract}
  Ca. 100 words.
  \end{abstract}
    \begin{keyword}
    reproducibility; regime shift; \sep 
    reproducibility; regime shift;
  \end{keyword}

 \end{frontmatter}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule
\endhead
C1 & Current code version & v0.9.0 \\
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & CC BY 4.0 \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
& & \\
& & \\
& & \\
& & \\
& & \\
C8 & If available Link to developer documentation/manual &
\textcolor{red}{TO BE ADDED} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Rainer@krugs.de \\
\bottomrule
\end{longtable}

@@

Mathematical models play a key role in the development of understanding
about how ecosystems work and how they respond to environmental changes
\citep{Binzer2016a, Chaparro-Pedraza2021, Vasseur2005}. They are also
critical for developing hypotheses that can then be tested in
experimental studies. One area of ecology in which simple models have
played a very influential role is the area of how ecosystems respond to
gradual environmental change \citep{REF_NEEDED}. - Scheffer, M.,
Carpenter, S., Foley, J. a, Folke, C. \& Walker, B. (2001). Catastrophic
shifts in ecosystems. Nature, 413, 591--6.

It is conceivable that an ecosystem state, such as the total biomass of
a particular type of bacteria, may remain unchanged when something about
its environment changes, such as the temperature. It is possible that
the ecosystem state changes gradually itself. It is also possible that
the ecosystem state may change abruptly into a new state that is
difficult to recover from \citep{Scheffer2001}.

This possibility for abrupt, perhaps catastrophic changes that are
difficult to reverse causes considerable concern
\citep{Collins2021, Northrop2021, Vandermeer2019}.

An example where the gradual change of a single environmental variable
is causing an abrupt change of the system is the switch from an aerobic
(oxygen is available for metabolism) to anaerobic (oxygen generally
unavailable) system. This system has been investigated by
\citet{Bush2017} in a simulation study of a mathematical model of the
system. They showed that gradual change in the environment could cause
catastrophic changes in the ecosystem state that would be difficult to
reverse.

One limitation of the study by \citet{Bush2017} was limited biodiversity
in the ecosystem model. Specifically, there were three functional groups
of bacteria, but within each of these groups, there was no biodiversity.
This then leaves open the question of if and how biodiversity within
functional groups, in their model, affects the ecosystem response to
environmental change. This limitation is not limited to the study of
\citet{Bush2017}, there are few if any studies of the effects of
biodiversity on abrupt transitions between ecosystem states.

We decided to fill this research gap, by making a simulation study of
how within functional group biodiversity affects ecosystem responses to
environmental change \citet{Limberger2022}, and to base our work on the
work and model of \citet{Bush2017}. It was with this goal in mind that
we developed the \emph{microxanox} package. The first stage of
development was to write code from scratch (as there was no available
code to start from) and to confirm that this new implementation would
reproduce the previously published results. The resulting reproduction
is available as one of the package vignettes:
\href{LINK\%20NEEDED}{vignette Partial reproduction of Bush et al}.

@@

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
biogeochemical processes and is based on \citet{Bush2017} (See
\citet{Bush2017} for a detailed discussion of the model). At the core of
the simulations is a set of ordinary differential equations (specified
in the function \texttt{bushplus\_dynamic\_model()}, though users do not
need to call this function directly). There are functions for running
individual simulations and for running a set of simulations across, for
example, a range of environmental conditions.

To make the simulation run with multiple species, we expressed different
species characteristics in the elements of vectors and matrices. We also
adapted the code for the ordinary differential equations to include the
vectors and matrices, and to use matrix mathematics. In this way, we
made it possible to run simulations with different numbers of species
without having to change the underlying code.

The package functions and code are structured in modular, so that some
new functionality can be easily added. E.g. the event definition can
easily be changed, or other aspects can be adjusted. All values in the
parameter object can be changed as needed and the general structure of
the code should make it not to difficult to adapt the model to other
similar systems.

@@

The framework used when writing this package aimed to maximise
simplicity for the user, and to make it easy to reproduce results (see
the supplement to \citet{Limberger2022} for an example of how this is
used). As such, all the parameters needed to run a simulation or find a
stable state are contained in a single object (which can easily be
created using included functions). This parameter object is given to a
function that runs the simulations and returns the results. The returned
results object is identical to the parameter object but with an
additional slot named \texttt{results}, which contains the results of
the run. Thus the returned results object contains the simulation
conditions (parameters) as well as the rsults, and can be used to run
the simulation again. This promotes reproducibility and makes
incremental changes of individual parameters and re-running the
simulations straightforward.

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package \citep{Soetaert2010}. The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains
among other things the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed please see the
\emph{User Guide} which accompanies the package or is available at
\href{@LINK_NEEDED}{User Guide}

@@

\hypertarget{finding-steady-states}{%
\subsection{Finding Steady States}\label{finding-steady-states}}

The general approach used to find the stable state of the system with a
specific parameter set is to run the simulation for a long time and
record the final state. When one does this across a range of
environmental conditions, one discovers how the steady state of the
system responds to the environmental conditions. The package contains
functionality for finding steady states that correspond to values of one
environmental variable, namely the value of oxygen diffusivity.

Two methods for finding steady states are implemented. The first runs a
separate simulation for each combination of starting conditions and
oxygen diffusivity (let us term this the \emph{Replication method}).
This is the method used in the \citet{Bush2017} study. The second runs
only two simulations, with step-wise and slowly temporally increasing
oxygen diffusivity, and the other with step-wise and slowly decreasing
or decreasing oxygen diffusivity. During this temporal environmental
change, the state of the system was recorded just before change to a new
oxygen diffusivity (let us term this the \emph{Temporal method}). We
implemented two methods since there is no definitive best method, and in
order to check if results were sensitive to choice of method.

The replication method is implementad in the function
\texttt{run\_replication\_ssfind()} which takes a parameter object as
returned by the function \texttt{new\_replication\_ssfind\_parameter()}
and the number of cores for multithreading the simulation. As the
multithreading uses the package function \texttt{mclapply()} from the
package \texttt{parallel} \citep{RCoreTeam2022}, the multthreading only
works on Linux and Mac. It is planned to move to \texttt{parLapply()}
\citep{RCoreTeam2022} in a future release.

@@

\hypertarget{visualising-and-analysing-results}{%
\subsection{Visualising and analysing
results}\label{visualising-and-analysing-results}}

From the raw results returned, summary measures about how the ecoystem
stable states respond to environmental change can be extracted. The
function \texttt{get\_stability\_measures()} returns quantities such as
the amount of environmental change required for the system to abruptly
change to a different state.

@@

{\centering \includegraphics[width=350px]{figures/ug_three_strains_dynamics}

\caption{Plot of results of a simulation run using the function $plot_dynamics()$. Details can be found in the "User Guide" section "Three strains per functional group".}\label{fig:plot-dynamics}
\end{figure}

@@

The free and open source implementation and extension of the model used
in \citet{Bush2017} provides the means of reproducing the results
published while at the same time provides the means of doing unique,
innovative, and important investigations of how ecosystems respond to
environmental change. The design of the package code and functionality
is with reproducibility in mind: the combination of all parameters being
in a single parameter object as well as the return of the simulation as
a result object which inherits from the parameter object provides a
relatively easy to use framework to implement reproducible experiments.

While the package was not intended to provide a code which can be very
easily adapted to other types of ecosystems and types of organism, it
seems likely that the overall framework could be used as a basis for
models of other ecosystems and organisms. For example, all such models
would have parameters that differ among species and need to be described
in an object. All would need to run simulations and sets of simulations
across environmental conditions. Researchers want to model a new
ecosystem do not, therefore, have to start from scratch. Nevertheless,
depending on how different were the new ecosystem to model, the require
code changes could be quite significant.

We evidence the impact of the \emph{microxanox} package by describing
three use cases, each of which was only possible with the package and
that contains new results. The first two use cases are described in
detail in the User Guide and the Partial Reproduction Vignettes. The
third is taken from \citet{REF_NEEDED} for which this R package was
designed. All of these use cases can be expanded to larger numbers of
strains per functional group and variable values van be changed.

@@

The package contains functionality to make a specific pattern of
temporal change in the environmental condition oxygen diffusivity; this
functionality forms the basis of the temporal method for finding stable
states. A simple example of this functionality is given in the Partial
Reproduction vignette, which we briefly show here (Figure @ref(fig:uc1).
The example is composed of a single simulation, at the beginning of
which the system is in the oxic state with high abundance of
cyanobacteria. Oyxgen diffusivity is then slowly decreased and
eventually, around hour 30'000 the system switches to the anoxic state,
with high abundance of both sulfur bacteria types. The oxygen
diffusivity is then quickly increased and at around hour 38'000 the
system abruptly switches back to the oxic state.

@@

{\centering \includegraphics[width=350px]{./figures/gen_uc1_partrep_temporal_state_switching}

@@

\caption{The temporal dynamics of the ecosystem model when an environmental condition (here parameter *a*, the oxygen diffusivity) changes. Plot of the stable states of the simulation runs under different oxygen diffusivity.}\label{fig:uc1}
\end{figure}

The package also contains a function to extract summary features of
ecosystem responses to environmental change, such as the amount of
hysteresis displayed by the ecosystem. Hysteresis is a key feature of
ecosystem responses to environmental change, because it is related to
how difficult it can be to reserve the effects of environmental change
\citep{Scheffer2001}. The amount of hysteresis was measured here as the
extent of the environmental condition for which there were two stable
states. I.e. it was the extent of the environmental conditions for which
historical conditions play an important role in determining the current
system state (a definition of hysteresis). The results show that the
amount of hysteresis depends greatly on the combinations of organisms
present (Figure @ref(fig:uc2)). For example, with only the CB
(cyanobacteria) present, there was no hysteresis. In contrast, the
presence of both CB and SB (sulfate reducing bacteria) led to a large
amount of hysteresis. (These results are also given in the Partial
Reproduction vignette.)

\textbf{\citet{Rainer}\ldots{} the figure needs beautifying please.}
\textbf{\citet{Owen}: is this OK like this?}

{\centering \includegraphics[width=350px]{./figures/gen_uc2_user_guide_hysteresis}

\caption{The amount of hysteresis depends on the combination of types of organisms present.}\label{fig:uc2}
\end{figure}

@@

>>>>

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

@@

\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header
%% \documentclass[preprint,12pt, a4paper]{elsarticle}

@@

\begin{document}
\begin{frontmatter}

  \title{Microxanox: an R package for simulating an aquatic
\(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
    \author[University of Zürich]{Rainer M Krug\corref{Corresponding
Author}}
   \ead{Rainer.Krug@uzh.ch; Rainer@krugs.de} 
    \author[University of Zürich]{Owen L. Petchey}
   \ead{Owen.Petchey@ieu.uzh.ch} 
      \address[University of Zürich]{Department of Evolutionary Biology
and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  Ca. 100 words.
  \end{abstract}
   \begin{keyword} reproducibility; regime shift;\end{keyword}
 \end{frontmatter}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.05}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.41}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.54}}@{}}
\toprule
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Please fill
in this column} \\
\midrule
\endhead
C1 & Current code version & v0.9.0 \\
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & CC BY 4.0 \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
& & \\
& & \\
& & \\
& & \\
& & \\
C8 & If available Link to developer documentation/manual &
\textcolor{red}{TO BE ADDED} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Rainer@krugs.de \\
\bottomrule
\end{longtable}

@@

Mathematical models play a key role in the development of understanding
about how ecosystems work and how they respond to environmental changes
(\protect\hyperlink{ref-Binzer2016a}{Binzer et al., 2016};
\protect\hyperlink{ref-Chaparro-Pedraza2021}{Chaparro-Pedraza, 2021};
\protect\hyperlink{ref-Vasseur2005}{Vasseur and McCann, 2005}). They are
also important for developing hypotheses that can then be tested in
empirical studies. One area of ecology in which simple models have
played a influential role is the area of how ecosystems respond to
gradual change in an environmental driver
(\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). An
environmental driver is a environmental condition that affects an
ecosystem, but is assumed to not be affected by the ecosystem, such as
the rate of nutrient input into a lake.

It is conceivable that an ecosystem state, such as the total biomass of
a particular type of bacteria, may remain unchanged when an
environmental driver changes. It is also possible that the ecosystem
state changes gradually. It is also possible that the ecosystem state
changes abruptly to a new state that is difficult to recover from
(\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). This
possibility for abrupt, perhaps catastrophic changes that are difficult
to reverse causes considerable concern
(\protect\hyperlink{ref-Collins2021}{Collins et al., 2021};
\protect\hyperlink{ref-Northrop2021}{Northrop et al., 2021};
\protect\hyperlink{ref-Vandermeer2019}{Vandermeer and Perfecto, 2019}).

An example where a gradual change of an environmental variable causes an
abrupt change of the system is the switch from an aerobic (oxygen is
available for metabolism) to anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. Three types of microbes occur in the
model: cyanobacteria (CB) dominating the oxic state, and two types of
sulfur bacteria that dominate the anoxic state (sulfate reducing
bacteria (SB) and phototrophic sulfur bacteria (PB)). This system has
been investigated by \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) in a simulation study of a
mathematical model. They showed that gradual change in the rate at which
oxygen could diffuse into the ecosystem (termed the oxygen
diffusitivity) could cause catastrophic changes in the ecosystem state
that would be difficult to reverse.

One feature of the study by \protect\hyperlink{ref-Bush2017}{Bush et
al.} (\protect\hyperlink{ref-Bush2017}{2017}) was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves open the question of if and how biodiversity
within these types (i.e.~functional groups) of bacteria, in their model,
affects the ecosystem response to environmental change. This limitation
is not specific to the study of \protect\hyperlink{ref-Bush2017}{Bush et
al.} (\protect\hyperlink{ref-Bush2017}{2017}), there are few if any
studies of the effects of biodiversity on abrupt transitions between
ecosystem states.

We decided to fill this research gap by making a simulation study of how
within functional group biodiversity affects ecosystem responses to
environmental change \protect\hyperlink{ref-Limberger2022}{Limberger}
(\protect\hyperlink{ref-Limberger2022}{2022}), and to base our work on
the work and model of \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}). It was with this goal in mind
that we developed the \emph{microxanox} package. The first stage of
development was to write code from scratch (as there was no available
code to start from) and to confirm that this new implementation would
reproduce the previously published results. The resulting reproduction
is available as one of the package vignettes:
\href{LINK\%20NEEDED}{vignette Partial reproduction of Bush et al}.

@@

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
organisms and biogeochemical processes and is based on
\protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) (See
\protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) for a detailed discussion of
the model). At the core of the simulations is a set of ordinary
differential equations (specified in the function
\texttt{bushplus\_dynamic\_model()}, though this function need not be
directly called). There are functions for running individual simulations
and for running a set of simulations across, for example, a range of
environmental conditions.

To make the simulation run with multiple species, we expressed different
species characteristics in the elements of vectors and matrices. We also
coded the ordinary differential equations to include the vectors and
matrices, and to use matrix mathematics. In this way, we made it
possible to run simulations with different numbers of species without
having to change the underlying code.

The package functions and code have modular structure, so that new
functionality can be easily added. E.g. temporally defined events of any
type could be specified. Further, all parameter values required to run a
simulation are stored in one object. Lastly, the general structure of
the code should make it straightforward to adapt the model to other
similar systems (described in more detail in the Impact section).

@@

The framework used when writing this package aimed to maximise
simplicity for the user, and to make it easy to reproduce results (see
the supplement to \protect\hyperlink{ref-Limberger2022}{Limberger}
(\protect\hyperlink{ref-Limberger2022}{2022}) for an example of how this
is used). As such, all the parameters needed to run a simulation or find
a stable state (i.e.~the final state of the ecosystem) are contained in
a single object (which can easily be created using included functions).
This parameter object is given to a function that runs the simulations
and returns the results. The returned results object is identical to the
parameter object but with an additional slot named \texttt{results},
which contains the results of the run. Thus the returned results object
contains the simulation conditions (parameters) as well as the results,
and can be used to run the simulation again. This promotes
reproducibility and makes incremental changes of individual parameters
and re-running the simulations straightforward.

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package
(\protect\hyperlink{ref-Soetaert2010}{Soetaert et al., 2010}). The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains
among other things the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed please see the
\emph{User Guide} which accompanies the package or is available at
\href{@LINK_NEEDED}{User Guide}

@@

\hypertarget{finding-stable-states}{%
\subsection{Finding Stable States}\label{finding-stable-states}}

The general approach used to find the stable state of the system with a
specific parameter set is to run the simulation for a long time and
record the final state. When one does this across a range of
environmental conditions, one discovers how the steady state of the
system responds to the environmental conditions. The package contains
functionality for finding steady states that correspond to values of one
environmental driver, namely the value of oxygen diffusivity.

Two methods for finding steady states are implemented. The first runs a
separate simulation for each combination of starting conditions and
oxygen diffusivity (we term this the \emph{Replication method}). This is
the method used in the \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) study. The second runs two
simulations, with step-wise and slowly temporally increasing oxygen
diffusivity, and the other with step-wise and slowly decreasing or
decreasing oxygen diffusivity. During this temporal environmental
change, the state of the system was recorded just before change to a new
oxygen diffusivity (we term this the \emph{Temporal method}). We
implemented two methods since there is no definitive best method, and in
order to check if results were sensitive to choice of method.

The replication method is implemented in the function
\texttt{run\_replication\_ssfind()} which takes a parameter object as
returned by the function \texttt{new\_replication\_ssfind\_parameter()}
and the number of cores for multithreading the simulation. As the
multithreading uses the package function \texttt{mclapply()} from the
package \texttt{parallel} (\protect\hyperlink{ref-RCoreTeam2022}{R Core
Team, 2022}), the multthreading only works on Linux and Mac. It is
planned to move to \texttt{parLapply()}
(\protect\hyperlink{ref-RCoreTeam2022}{R Core Team, 2022}) in a future
release.

@@

\hypertarget{analysing-and-visualising-results}{%
\subsection{Analysing and visualising
results}\label{analysing-and-visualising-results}}

From the results returned, summary measures about how the ecosystem
stable states respond to environmental change can be extracted. The
function \texttt{get\_stability\_measures()} returns quantities such as
the amount of environmental change required for the system to abruptly
change to a different state.

@@

\textbackslash begin\{figure\}

\{\centering \includegraphics[width=350px]{figures/ug_three_strains_dynamics}

\}

\textbackslash caption\{Results of a simulation run shown using the
function \(plot_dynamics()\). In this case, there were three strains per
functional group, though strains within functional groups had identical
properties in this example. CB\_1 = cyanobacteria strain 1; SB\_1 =
sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria
strain 1. Here we show the figure produce by \(plot_dynamics()\), which
produces figures not with all features ideal for
publication.\}\label{fig:plot-dynamics} \textbackslash end\{figure\}

@@

The open source implementation and extension of the model used in
\protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) provides the means of
reproducing the results published while at the same time provides the
means of doing unique, innovative, and important investigations of how
ecosystems respond to environmental change, and how biodiversity may
modulate this response.

The design of the package code and functionality is with reproducibility
in mind: the combination of all parameters being in a single parameter
object as well as the return of the simulation as a result object which
inherits from the parameter object provides a relatively easy to use
framework to implement reproducible experiments.

Here we evidence the impact of the \emph{microxanox} package by
describing three use cases and then by describing how the package can be
a starting point for models of other ecosystems. The first two use cases
are described in detail (including the code for reproducing them) in the
\emph{User Guide} and the \emph{Partial Reproduction} vignettes. The
third is taken from \protect\hyperlink{ref-REF_NEEDED}{REF NEEDED}
(\protect\hyperlink{ref-REF_NEEDED}{2222}).

@@

The study of \protect\hyperlink{ref-Bush2017}{Bush et al.}
(\protect\hyperlink{ref-Bush2017}{2017}) includes simulations of the
effect of oxygen diffusivity (an environmental driver, in the sense that
it affects the ecosystem but is not affected by it) on the ecosystem
state (oxic or anoxic). The \emph{microxonax} package contains
functionality to make a specific temporal pattern of change in the
oxygen diffusivity. As well as allowing individual simulations during
which oxygen diffusivity varies, this functionality forms the basis of
the temporal method for finding stable states.

An example of this functionality is given in the \emph{Partial
Reproduction} vignette, which we briefly show here (Figure
@ref(fig:uc1). The example is composed of a single simulation, at the
beginning of which the system is in the oxic state with high abundance
of cyanobacteria. Oyxgen diffusivity is then slowly decreased and
eventually, around hour 30'000 the system switches to the anoxic state,
with high abundance of both sulfur bacteria types. The oxygen
diffusivity is then increased and at around hour 38'000 the system
abruptly switches back to the oxic state.

Also visible in the results are thick lines showing abundances of
bacteria when abundances are low. This is due to the implementation of a
function that at regular intervals, increases the abundance to a preset
level. This prevents abundances reducing to very small numbers. The
function that implements this increase abundance can also be made to add
a certain abundance to each strain at regular intervals, thus simulating
immigration in to the system.

\begin{verbatim}
#> 
#> Attaching package: 'dplyr'
#> The following objects are masked from 'package:stats':
#> 
#>     filter, lag
#> The following objects are masked from 'package:base':
#> 
#>     intersect, setdiff, setequal, union
\end{verbatim}

\textbackslash begin\{figure\}

\{\centering \includegraphics[width=350px]{/Users/owenpetchey/Desktop/microxanox/microxanox/inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching}

\}

\textbackslash caption\{The temporal dynamics of the ecosystem model
when an environmental condition (here parameter \emph{a}, the oxygen
diffusivity) changes. Plot of the stable states of the simulation runs
under different oxygen diffusivity. In this simulation there is only one
strain in each functional group. CB\_1 = cyanobacteria strain 1; SB\_1 =
sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria
strain 1. Here we show a figure adapted from the output of the
\(plot_dynamics()\) function.\}\label{fig:uc1}
\textbackslash end\{figure\}

The package contains a function to extract summary features of ecosystem
responses to environmental change, such as the amount of hysteresis
displayed by the ecosystem. Hysteresis is a key feature of ecosystem
responses to environmental change, because it is related to how
difficult it can be to reserve the effects of environmental change
(\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). The
amount of hysteresis is measured as the extent of the environmental
condition (here oxygen diffusitivity) for which there were two stable
states. I.e. it is the extent of the environmental conditions for which
historical conditions play an important role in determining the current
system state (a definition of hysteresis).

Using the package to calculate the extent of hysteresis involves setting
ecosystem and simulation parameters, including parameters for the
finding of stable states across an environmental gradient, running the
stable state finding function, and analysing the results with the
function that calculates extent of hysteresis. The code for this is
provided in the \emph{User Guide} vignette.

The results show that the amount of hysteresis depends greatly on the
combinations of organisms present (Figure @ref(fig:uc2)). For example,
with only the CB (cyanobacteria) present, there was no hysteresis. In
contrast, the presence of both CB and SB (sulfate reducing bacteria) led
to a large amount of hysteresis. (These results are also given in the
\emph{Partial Reproduction} vignette.)

@@

{\centering \includegraphics[width=150px]{/Users/owenpetchey/Desktop/microxanox/microxanox/inst/manuscript/figures/gen_uc2_user_guide_hysteresis}

@@

\caption{The amount of hysteresis depends on the combination of types of organisms present. The model is entirely deterministic, hence there are no error bars.}\label{fig:uc2}
\end{figure}

@@

\hypertarget{adapting-for-other-ecosystems-and-organisms}{%
\subsection{Adapting for other ecosystems and
organisms}\label{adapting-for-other-ecosystems-and-organisms}}
